---
title: "Reproducible Research: Peer Assessment 1"
output: 
    html_document:
        keep_md: true
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
    fig.width=6, 
    fig.height=4, 
    fig.path='figure/',
    echo=TRUE, 
    warning=FALSE, 
    message=FALSE
)
```


## Loading and preprocessing the data

The data is loaded into a variable called **activity**, assuming that the file is called "activity.csv" and is in a subfolder of the current working directory called "data".

``` {r}

activity <- read.csv("data/activity.csv")

```

The date column is converted from type factor to type date to help with processing later.

``` {r}

library(dplyr)
activity <- mutate(activity, date = as.Date(date))

```
  
  
  
## What is mean total number of steps taken per day?

Use group_by and summarise functions to obtain the total number of steps per day:

``` {r}

activityByDay <- group_by(activity, date) %>% 
                summarise(total = sum(steps, na.rm=TRUE))

```


The following histogram shows the frequency of total number of steps per day:

``` {r plotFrequencyTotalSteps}
library(ggplot2)

ggplot(
        data=activityByDay, 
        aes(total)
    ) +
    geom_histogram(
        col="darkgreen",
        fill="palegreen3",
        binwidth=1000
    ) + 
    labs(
        x="Total Steps per Day", 
        y="Frequency"
    )

```


The mean and median of the total number of steps taken per day are calculated as follows:

``` {r}

meanTotal <- round(mean(activityByDay$total), digits = 0)
medianTotal <- median(activityByDay$total)

```

This gives the mean as `r meanTotal` and the median as `r medianTotal`.  



## What is the average daily activity pattern?

The data is grouped by each 5 minute interval, and the average for each interval is calculated.  This is then plotted in the following graph:

``` {r plotMeanForIntervals}
activityByInterval <- group_by(activity, interval) %>% 
                        summarise(average = mean(steps, na.rm=TRUE))

ggplot(
        data=activityByInterval, 
        aes(interval, average)
    ) + 
    geom_line(
        col="darkgreen"
    ) + 
    labs(
        x="Interval", 
        y="Average across all days"
    )

```


The 5-minute interval, which on average across all the days in the dataset, contains the maximum number of steps is calculated as follows

``` {r}

maxSteps <- filter(activityByInterval, average == max(activityByInterval$average))$interval

```

This gives the interval `r maxSteps`.



## Imputing missing values

The total number missing values is calculated as follows:

``` {r}

totalNA <- filter(activity, is.na(steps)) %>% summarise(total = n())

```

This gives the total number if missing values as `r totalNA`.


The missing values can be replaced using the average value for that 5 minute interval across the whole period.  This should be rounded to the nearest integer as the number steps should not contain decimals.

Using this strategy, a new dataset called **activityFixNA** is created with no missing values:


``` {r}

activityFixNA <- inner_join(
                    activity, 
                    activityByInterval, 
                    by=c("interval" = "interval")
                ) %>% 
                mutate(
                    steps = ifelse(
                        is.na(steps), 
                        round(average, digits=0), 
                        steps
                    )
                ) %>% 
                select(steps, date, interval)

```


If we group by date, the following histogram shows the frequency of total number of steps per day, now that the missing values have been removed:

``` {r plotMeanForIntervalsFixNA}

activityByDayFixNA <- group_by(activityFixNA, date) %>% 
                        summarise(total = sum(steps))

ggplot(
        data=activityByDayFixNA, 
        aes(activityByDayFixNA$total)
    ) +
    geom_histogram(
        col="darkgreen",
        fill="palegreen3",
        binwidth=1000
    ) + 
    labs(
        x="Total Steps per Day", 
        y="Frequency"
    )

```


The new mean and median are calculated as follows:

``` {r}

meanTotalFixNA <- round(mean(activityByDayFixNA$total), digits = 0)
medianTotalFixNA <- median(activityByDayFixNA$total)

```

This gives the mean as `r as.character(meanTotalFixNA)` and the median as `r as.character(medianTotalFixNA)`.  

By imputing the missing data, both the mean and median values are calculated as higher numbers and the difference is smaller.




## Are there differences in activity patterns between weekdays and weekends?

A new factor variable **dayType** is created to split the data between weekday and weekend:

``` {r}

activityFixNA <- mutate(
                    activityFixNA, 
                    dayType = factor(
                        ifelse(
                            weekdays(date)=="Saturday" | weekdays(date)=="Sunday", 
                            "weekend", 
                            "weekday"
                        )
                    )
                )

```


The following panel plot shows the difference in the average number of steps per 5 minute interval between weekdays and weekends:

``` {r plotCompareDayType,fig.height=8}
activityByIntervalFixNA <- group_by(activityFixNA, interval, dayType) %>% 
                            summarise(average = mean(steps))

library(lattice)

xyplot(
    average ~ interval | dayType, 
    data=activityByIntervalFixNA, 
    type="l", 
    xlab="Interval", 
    ylab="Number of steps", 
    layout=c(1,2), 
    col="darkgreen"
)

```
